Imprecise Datapath Design: An Overclocking Approach

### 2013-05-28 TRETS2013/TRETS_overclocking.tex
Add Contribution

\category{C.2.2}{Computer-Communication Networks}{Network Protocols}

\terms{Design, Algorithms, Performance}

\keywords{A,B}

@@

\begin{bottomstuff}
This work is supported by the EPSRC, under
grant EP/I020557/1 and grand EP/I012036/1.

@@

\section{Introduction}
Circuit performance has increased tremendously over the past decades, with the continuous scaling of CMOS technology to the nanometer regime. However, the drastic variations introduced by higher integration densities is anticipated to be the major obstacle when designing reliable, high performance circuits. Currently, hardware designers tend to employ conservative safety margins to ensure a uniform functionality across a variety of possible working environments. However, continuing with this approach could become very costly in the short term future. This is because the highly scaled CMOS devices would inevitably exhibit probabilistic or statistical behavior [xxx], meaning that we can no longer expect a uniform circuit performance. Furthermore, covering all possible worst cases would become increasingly difficult, expensive and result in large yield loss because of the variation [xxx].

@@

In order to boost the operating frequency of a datapath, the standard approaches are either to heavily pipeline the design or to reduce the precision throughout the datapath. For the former method, it should be noted that pipelining will not tend to reduce the circuit latency because the latency in terms of clock cycles will increase. As a result, this method will not be applicable to many embedded applications, which typically have strict latency requirements, or in any datapath containing feedback where C-slow retiming is inappropriate. For the second approach, reducing the datapath precision would reduce the latency at the cost of introducing quantization errors into the design. This has inspired research into exploiting the potential benefits of using the minimum precision necessary to satisfy a design specification~\cite{GeorgeDT11}. A brief review of existing approaches in this area will be discussed in Section~\ref{Section_Background_WL}.

@@

In this paper, we describe an alternative circuit design methodology when considering tradeoffs between accuracy, performance and area. We suggest that for certain applications it is beneficial to move away from the traditional model of creating a conservative design that is guaranteed to avoid timing violations. Instead, it may be preferable to create a design in which timing violations may occur, under the knowledge that they only occur rarely because specific input patterns are required to generate the worst case errors. This paper elaborates on the prior work [xxx], by incorporating silicon area as another evaluation metric. This enables us to find the optimum design choice of the basic arithmetic operators (i.e. binary adders) under various design constraints. A summary of the main contributions of this work are as follows:

The rest of this paper is organized as follows. It first reviews in detail the current literature regarding the approaches used in word-length optimization and approximate datapath design in Section~\ref{Section_Background}. The analytical error models for two different adder structures: RCA and CSA, are discussed in Section~\ref{section_RCA} and Section~\ref{Section_CSA}, respectively. The decision of the optimum adder structure under the accuracy-performance-area trade-offs, is then put forward in Section~\ref{Section_OptimumAdder}. It is followed by a description of the probabilistic error models for CCM in Section~\ref{section_CCM}. Section~\ref{Section_ExpSetup} discusses a practical experimental setup to verify our models and test the proposed design methodology. We then demonstration the benefits of our proposed approach in Section~\ref{Section_Experiments}. Finally, Section~\ref{Section_Conclusion} discusses the conclusions of this work and the possible research directions in the future.

\section{Background}\label{Section_Background}
\subsection{Word-length Optimization}\label{Section_Background_WL}
There exists a significant amount of work demonstrating that optimizing the precision used throughout a datapath would bring a substantial benefits on clock speed, silicon area and power consumption. There are two main research directions in this area. The first discusses methods to analyze the errors caused by the usage of finite precision and compute bounds on the worst-case errors seen at the output of the design. Related techniques include the simulation-based approach~\cite{SungWLop_Simulation} and the analytical-based approach such as interval arithmetic~\cite{moore1966IA}, affine arithmetic~\cite{de2004affine}, Satisfiability-Modulo Theories~\cite{SAT2010} and polynomial representations~\cite{Boland2011TCAD}. A detailed summary of these methods can be found in~\cite{GeorgeDT11,lee2006accuracy}. The second stream uses these tools to choose the minimum precision within a datapath to satisfy an required accuracy specification~\cite{RoldaoWLop_Simulation}. In this work, we speculate and employ the maximum word-length in the datapath under a given timing constraint while ensuring no timing errors will occur. The truncation errors derived in this process are then compared to the overclocking errors generated in our proposed new scenario.

\subsection{Imprecise Circuit Design Methodology}\label{Section_Background_Imprecise}
However, the choice of precision is not the only source of error when designing a datapath. Recently we have seen a growth of parallel streams of research which aim at exploring alternative methods to trade accuracy for design efficiency. This strand of research is motivated by the fact that extra benefits of manufacturing, test, power and timing can be obtained if the tight requirement of absolute correctness is released for devices and interconnects, as pointed out by the international technology roadmap for semiconductors (ITRS) in 2007~\cite{semiconductor2007international}. This topic is expected to be of growing importance in the future, because we will face new challenges as the technology scales further.

According to the existing literatures, one way to operating circuits beyond the deterministic region is to relax the design constraints and the guard-bands that are conventionally used to avoid the worst cases. A series of work known as ``Better-Than-Worst-Case (BTWC) Design'' introduced a universal framework to push the circuit performance to its limits~\cite{BetterThanWS2005}. In general, a BTWC design is composed of cores and checkers. The cores are operated with high performance by eliminating the guard-bands, meanwhile the possible timing errors are diagnosed by the checkers. Optionally the system can be recovered at the observation of errors. As an exemplary design, the Razor project~\cite{Razor2003,Razor2004} was proposed to shave the conservative timing margins by overscaling the supply voltage and clock frequency, while monitoring the output error rates by utilizing a self-checking circuit. This work demonstrated that the benefits brought by removing the safe margin outweigh the cost of monitoring and recovering from errors. For example, $22\%$ or over $30\%$ power consumption can be saved with $\sim0.01\%$ or $\sim1\%$ error rates at the output, respectively. Related work includes a similar frequency overscaling technique by operating circuits slightly slower than the critical path delay with dedicated checker circuits to ensure that timing errors will not occur~\cite{TEAtime2004}, or developing timing analysis tools that decide the optimum operating frequencies in the non-deterministic region due to process variation~\cite{ProbabilisticSTA}. However, the major problem of the BTWC work is the extra cost of silicon area due to the implementation of the checker circuits.

Another field of study has shown that errors can be potentially tolerated in various applications, e.g. DSP applications concerning human perceptions. In this case, even greater performance/power benefits can be achieved, as the design overhead of the checkers can be eliminated. Current research in this area focuses on designing probabilistic or imprecise circuits both at a software level and a hardware level. For instance at the programming language level, a tool was proposed~\cite{EnerJ2011Uwash,Truffle2012Uwash} to divide the program into the precise parts and the approximate parts, both of which are mapped to different hardware with different speed-grades, supply voltages etc. This technique enables a relatively high service quality, in the meantime energy reduction can be obtained due to approximation. Similar ideas have also been applied on hardware directly. As an example, Palem et~al.~\cite{NonUniformScaling} described a non-uniform voltage scaling technique for the ripple carry adder. In this study, multiple voltage regions are employed for different bits along a carry chain. That is, higher voltage would be applied for computations generating most significant bits, and vice versa. Nevertheless, utilizing several voltage regions within a ripple carry adder will rarely be practical in real situations. Furthermore, the overhead of applying this technique is not addressed.

While the aforementioned literatures take advantage of the fact that only specific input patterns could cause timing errors, research on imprecise architectures take this one step further by designing simplified circuits for peformance/energy efficiency. For example, Lu et~al. proposed a ``shrinking'' datapath that can be utilized to mimic and speculate the original logic functions~\cite{IntelSpeeding}. Kulkarni et.al. described an underdesigned $2\times2$ multiplier unit, of which the worst case was replaced by a normal case based on the straight-forward Karnaugh-Map analysis~\cite{Undersigned2x2multiplier}. In both cases, reduction of area and power consumption can be achieved with the cost of accuracy. In addition, Gupta et~al. developed approximate adders at the transistor level and compared the energy efficiency of their proposed architectures over truncation of input word-length of conventional structures~\cite{Gupta2013TransCADICS}. However, we should note that using imprecise architectures means correct results would never be obtained for certain input patterns. Another limitation is that the link between the probability of output correctness and energy savings or performance improvements are rarely analyzed from current research. In addition, these techniques cannot be directly applied onto existing hardware platforms such as FPGAs.

In this work, we propose an alternative methodology to remove the limitations of those existing approaches. It is argued in this paper that our method can be easily applied onto existing hardware arithmetic operators such as RCAs and CCMs with almost no hardware overhead. We support this hypothesis experimentally and analytically by presenting theoretical probabilistic error models.

\section{Ripple Carry Adder}\label{section_RCA}
\subsection{Adder Structures in FPGAs}
Adders serve as a key building block for arithmetic operations. Generally speaking, the ripple carry adder (RCA) is the most straightforward and widely used adder structure. As such, the philosophy of our approach is first exemplified with the analysis of a RCA. We later describe how this methodology can be extended to other arithmetic operators in Section~\ref{section_CCM} by discussing  the CCM that is commonly used in DSP applications and numerical algorithms.

Typically the maximum frequency of a RCA is determined by the longest carry propagation. Consequently, modern FPGAs offer built-in architectures for very fast ripple carry addition. For instance, the Altera Cyclone series uses fast tables~\cite{AlteraCyclone} while the Xilinx Virtex series employs dedicated multiplexers and encoders for the fast carry logic~\cite{Virtex6}. Fig.~\ref{FPGA adder} illustrates the structure of an $n$-bit RCA, which is composed of $n$ serial-connected full adders (FAs) and utilizes the internal fast carry logic of the Virtex-6 FPGA.

While the fast carry logic reduces the time of each individual carry-propagation delay, the overall delay of carry-propagation will eventually overwhelm the delay of sum generation of each LUT with increasing operand word-lengths. For our initial analysis, we assume that the carry propagation delay of each FA is a constant value $\mu$, which is a combination of logic delay and routing delay, and hence the critical path delay of the RCA is $\mu_{RCA}=n\mu$, as shown in~Fig.~\ref{FPGA adder}. For an $n$-bit RCA, it follows that if the sampling period $T_S$ is greater than $\mu_{RCA}$, correct results will be sampled. If, however, $T_S<\mu_{RCA}$, intermediate results will be sampled, potentially generating errors.

@@

\begin{table*}[htbp]
 %\vspace{-2ex}
 %% increase table row spacing, adjust to taste
   \renewcommand{\arraystretch}{1.05}
   \setlength{\tabcolsep}{3.5pt}
   \tbl{Relative Reduction of MRE ($\%$) in Overclocking Scenario for Various Normalized Frequencies Based on~(\ref{Relative Reduction MRE}) for two types of input data: the uniform data (Uni) and real image data from Lena (Lena).}{
   \centering
   \footnotesize
   %\vspace{-4mm}
   \begin{threeparttable}
   \begin{tabular}{|c|cc|cc|cc|cc|cc||cc|}
     \hline
     \multirow{2}*{\begin{tabular}{c}\textbf{Normalized}\\\textbf{Frequency}\end{tabular}} &
     \multicolumn{2}{c|}{\textbf{FIR}} &
     \multicolumn{2}{c|}{\textbf{Sobel}} &
     \multicolumn{2}{c|}{\textbf{IIR}} &
     \multicolumn{2}{c|}{\textbf{Butterworth}} &
     \multicolumn{2}{c||}{\textbf{DCT4}} &
     \multicolumn{2}{c|}{\textbf{Geo.Mean}}\\
     &  {Uni} & {Lena } &  {Uni} & {Lena } &  {Uni} & {Lena }
     &  {Uni} & {Lena } &  {Uni} & {Lena } &  {Uni} & {Lena }\\
     \hline
     1.04 & 99.85	& 100.00  & 99.51 &	99.74 &	 72.28 &	90.09 &	 79.03 &	 100.00 &	 83.58 &	 98.06 & 86.14 & 97.50\\
     1.08 & 98.93	& 99.97   & 96.26 &	93.75 &	 71.64 &	90.50 &	 78.81 &	 100.00 &	 83.26 &	 98.45 & 85.15 & 96.46\\
     1.12 & 94.27	& 98.82   & 96.25 &	93.62 &	 73.63 &   88.25 &	 81.88 &	 84.87  &	 89.44 &	 99.56 & 86.68 & 92.84\\
     1.16 & 99.66	& 99.91   & 73.73 &	93.62 &	 73.10 &   89.92 &	 79.30 &	 84.23  &	 79.07 &	 99.44 & 80.44 & 93.23\\
     1.20 & 96.03	& 99.90   & 81.55 &	81.52 &	 70.76 &	75.67 &	 64.96 &	 84.50  &	 N/A\tnote{*}	  & N/A\tnote{*} & 77.46 & 84.95 \\
     1.24 & 98.46	& 99.32   & 81.43 &	81.67 &	 70.47 &	76.12 &	 63.66 &	 84.23  &	 N/A\tnote{*}	  & N/A\tnote{*} & 77.44 & 84.92\\
     1.28 & 95.39	& 99.29   & 60.41 &	78.24 &	 N/A\tnote{*}	    &   N/A\tnote{*}     &	54.38 &	 75.15  &	N/A\tnote{*}	  & N/A\tnote{*} & 67.92 & 83.58\\
     1.32 & 95.37	& 98.75   & N/A\tnote{*}	  &	 N/A\tnote{*}	    &	 N/A\tnote{*}	    &	 N/A\tnote{*}   & N/A\tnote{*}   & N/A\tnote{*}    &    N/A\tnote{*}   & N/A\tnote{*} & 95.37	& 98.75 \\
     \hline
   \end{tabular}
   \normalsize
   \begin{tablenotes}
   \footnotesize
    \item[*] Current frequency cannot be achieved in the traditional scenario. These points are excluded from the calculation of geometric means.
   \end{tablenotes}
   \end{threeparttable}}
   \label{Ratio_MRE}
% \vspace{-1ex}
 \end{table*}

\begin{table*}[htbp]
 %\vspace{-2ex}
 %% increase table row spacing, adjust to taste
   \renewcommand{\arraystretch}{1.05}
   \setlength{\tabcolsep}{3.5pt}
   \tbl{Frequency Speedups ($\%$) in Overclocking Scenario Under Various Error Budgets for two types of input data: uniform data (Uni) and real image data from Lena (Lena).}{
%   \vspace{-2ex}
   \label{Exp_Real_Inputs}
   \centering
   \footnotesize
   \begin{tabular}{|c|cc|cc|cc|cc|cc||cc|}
     \hline
     \multirow{2}*{\begin{tabular}{c}\textbf{Error}\\\textbf{Budget($\%$)}\end{tabular}} &
     \multicolumn{2}{c|}{\textbf{FIR}} &
     \multicolumn{2}{c|}{\textbf{Sobel}} &
     \multicolumn{2}{c|}{\textbf{IIR}} &
     \multicolumn{2}{c|}{\textbf{Butterworth}} &
     \multicolumn{2}{c||}{\textbf{DCT4}} &
     \multicolumn{2}{c|}{\textbf{Geo.Mean}} \\
     %\multirow{2}*{\begin{tabular}{c}\textbf{Geo.}\\\textbf{Mean}\end{tabular}}\\
     & {Uni} & {Lena } &  {Uni} & {Lena } &  {Uni} & {Lena }
     & {Uni} & {Lena } &  {Uni} & {Lena } &  {Uni} & {Lena }\\
     \hline
     0.05 & 4.76 &	21.43&	 6.82&	6.82&	0.95&	 1.26&	12.40&	 24.03&	 0.72&	 0.96  &3.07 & 5.32\\
     0.5  & 19.05&	21.43&	 13.64&	6.82&	0.63&	 10.06&	24.03&	 24.03&	0.48&	 12.44   &4.52 &13.45\\
     1    & 19.05&	28.57&	 13.64&	18.18&	 10.06&	16.35&	 24.03&	 24.03&	 0.48&	 12.44    &7.86 &19.10\\
     5    & 19.15&	25.53&	 18.18&	18.18&	 0.54&	0.82&	 0.63&	 0.94&	 7.66&	 12.44    &3.91    &5.36\\
     10   & 19.15&	25.53&	 10.64&	10.64&	 0.54&	1.09&	 6.92&	 13.21&	 4.91&	 4.911  &5.19  &7.19\\
     20   & 19.15&	25.53&	 5.77&	15.39&	 8.70&	8.70&	 3.26&	 3.26&	 6.70&	 10.88    &7.32    &10.39\\
     50   & 15.69&	15.69&	 10.53&	19.30&	 42.50&	50.00&	 46.74&	 54.89&	 15.06&	 19.25    &21.8 &27.59\\
     \hline
   \end{tabular}}
   \normalsize
   \label{Max_Frequency}
%   \vspace{-3ex}
 \end{table*}

>>>>

\category{C.2.2}{Category put here}{Detail category}

\terms{Design, FPGA, Performance}

\keywords{Overclocking, Imprecise Design}

@@

\begin{bottomstuff}
This work is supported by the EPSRC, under
grant EP/I020557/1 and grant EP/I012036/1.

@@

\section{Introduction}
With the continuous scaling of CMOS technology to the nanometer regime, circuit performance has increased tremendously over the past decades. However, the drastic variations introduced by higher integration densities is anticipated to be the major obstacle when designing reliable, high performance circuits. Currently, hardware designers tend to employ conservative safety margins to ensure a uniform functionality across a variety of possible working environments. Nevertheless, continuing with this approach could become very costly in the short term future. This is because the highly scaled CMOS devices would inevitably exhibit probabilistic or statistical behavior, meaning that we can no longer expect a uniform circuit performance. Furthermore, covering all possible worst cases would become increasingly difficult, expensive and result in large yield loss because of the variation.

@@

In order to boost the operating frequency of a datapath, the standard approaches are either to heavily pipeline the design or to reduce the precision throughout the datapath. For the former method, it should be noted that pipelining will not tend to reduce the latency of the datapath, whilst the latency in terms of clock cycles will actually increase. As a result, this method will not be applicable to many embedded applications, which typically have strict latency requirements, or in any datapath containing feedback where C-slow retiming is inappropriate. For the second approach, reducing the datapath precision would reduce the latency at the cost of introducing quantization errors into the design. This has inspired research into exploiting the potential benefits of using the minimum precision necessary to satisfy a design specification~\cite{GeorgeDT11}. A brief review of existing approaches in this area will be discussed in Section~\ref{Section_Background_WL}.

@@

In this paper, we describe an alternative circuit design methodology when considering trade-offs between accuracy, performance and silicon area. We suggest that for certain applications it is beneficial to move away from the traditional model of creating a conservative design that is guaranteed to avoid timing violations. Instead, it may be preferable to create a design in which timing violations may occur, under the knowledge that they only occur rarely because specific input patterns are required to generate the worst case errors. This paper elaborates on the prior work [xxx], by incorporating silicon area as another evaluation metric. This enables us to find the optimum design choice of the basic arithmetic operators (i.e. binary adders) under various design constraints. A summary of the main contributions of this work are as follows:

\begin{itemize}
	\item Detailed modeling methods of probabilistic errors generated in basic arithmetic primitives from two design scenarios to trade accuracy for performance: overclocking the datapath or truncating precisions that used throughout the datpath,
	\item Exploring the accuracy-performance-area trade-offs for key arithmetic operators to illustrate the optimum architecture for a certain arithmetic function under a given requirement of silicon area,
	\item Demonstration of the proposed findings analytically and experimentially from FPGA implementations that allowing rare timing violations to happen leads to an reduction of average errors or an improvement of performance over trancating the datapath precisions to meet timing.
\end{itemize}

The rest of this paper is organized as follows. It first reviews in detail the current literature regarding the approaches used in word-length optimization and approximate datapath design in Section~\ref{Section_Background}. The analytical error models for two different adder structures: the ripple carry adder (RCA) and the carry-select adder (CSA), are discussed in Section~\ref{section_RCA} and Section~\ref{Section_CSA}, respectively. The decision of the optimum adder structure under the accuracy-performance-area trade-offs, is then put forward in Section~\ref{Section_OptimumAdder}. It is followed by a description of the probabilistic error models for another arithmetic operator, the constant coefficient multiplier (CCM), in Section~\ref{section_CCM}. Section~\ref{Section_ExpSetup} discusses a practical experimental setup to verify our models and test the proposed design methodology. We then demonstrate the benefits of our proposed approach in Section~\ref{Section_Experiments}. Finally, conclusion of this work and possible future work are discussed in Section~\ref{Section_Conclusion}. 
%discusses the conclusions of this work and the possible research directions in the future.

\section{Background}\label{Section_Background}
\subsection{Word-length Optimization}\label{Section_Background_WL}
There exists a significant amount of work demonstrating that optimizing the precision used throughout a datapath would bring a substantial benefits on clock speed, silicon area and power consumption. There are two main research directions in this area. The first discusses methods to analyze the errors caused by the usage of finite precision and compute bounds on the worst-case errors seen at the output of the design. Related techniques include the simulation-based approach~\cite{SungWLop_Simulation} and the analytical-based approach such as interval arithmetic~\cite{moore1966IA}, affine arithmetic~\cite{de2004affine}, Satisfiability-Modulo Theories~\cite{SAT2010} and polynomial representations~\cite{Boland2011TCAD}. A detailed summary of these methods can be found in~\cite{GeorgeDT11,lee2006accuracy}. In the second research field, these tools are utilized to choose the minimum precision within a datapath to satisfy a required accuracy specification~\cite{RoldaoWLop_Simulation}. In this work, we speculate and employ the maximum word-length in the datapath under a given timing constraint while ensuring no timing errors will occur. The truncation errors derived in this process are then compared to the overclocking errors generated in our proposed new scenario.

\subsection{Imprecise Circuit Design Methodology}\label{Section_Background_Imprecise}
It is worth noting that the choice of precision is not the only source of error when designing a datapath. Recently we have seen a growth of parallel streams of research which aim at exploring alternative methods to trade accuracy for design efficiency. This strand of research is motivated by the fact that extra benefits of manufacturing, test, power and timing can be obtained if the tight requirement of absolute correctness is released for devices and interconnects, as pointed out by the International Technology Roadmap for Semiconductors (ITRS) in 2007~\cite{semiconductor2007international}. This topic is expected to be of growing importance in the future, because we will face new design challenges as the technology scales further.

According to the existing literatures, one way to operating circuits beyond the deterministic region is to relax the design constraints and the guard-bands that are conventionally used to avoid the worst cases. A series of work known as ``Better-Than-Worst-Case (BTWC) Design'' introduced a universal framework to push the circuit performance to its limits~\cite{BetterThanWS2005}. In general, a BTWC design is composed of cores and checkers. The cores are operated with high performance by eliminating the guard-bands, meanwhile the possible timing errors are diagnosed by the checkers. Optionally the system can be recovered at the observation of errors. As an exemplary design, the Razor project~\cite{Razor2003,Razor2004} was proposed to shave the conservative timing margins by overscaling the supply voltage and clock frequency, while monitoring the output error rates by utilizing a self-checking circuit. This work demonstrated that the benefits brought by removing the safe margin outweigh the cost of monitoring and recovering from errors. For example, $22\%$ or over $30\%$ power consumption can be saved with $\sim0.01\%$ or $\sim1\%$ error rates at the output, respectively. Related work also includes a similar frequency overscaling technique by operating circuits slightly slower than the critical path delay with dedicated checker circuits to ensure that timing errors will not occur~\cite{TEAtime2004}, or developing timing analysis tools that decide the optimum operating frequencies in the non-deterministic region due to process variation~\cite{ProbabilisticSTA}. However, the major problem of the BTWC work is the extra cost introduced by implementing the checker circuits.
%of silicon area due to the implementation of the checker circuits.

Another field of study has shown that errors can be potentially tolerated in various applications, e.g. DSP applications concerning human perceptions. In this case, even greater performance or power benefits can be achieved, as the design overhead of the checkers can be eliminated. Current research in this area focuses on designing probabilistic or imprecise circuits both at a software level and a hardware level. For instance at the programming language level, a tool was proposed~\cite{EnerJ2011Uwash,Truffle2012Uwash} to divide the program into the precise parts and the approximate parts, which are mapped to different hardware with different speed-grades, supply voltages etc., respectively. This technique enables a relatively high service quality, in the meantime energy reduction can be obtained due to approximation. Similar ideas have also been applied on hardware directly. As an example, Palem et~al.~\cite{NonUniformScaling} described a non-uniform voltage scaling technique for the ripple carry adder. In this study, multiple voltage regions are employed for different bits along a carry chain. That is, higher voltage would be applied for computations generating most significant bits, and vice versa. Nevertheless, utilizing several voltage regions within a ripple carry adder will rarely be practical in real situations. Furthermore, the overhead of applying this technique is not addressed.

While the aforementioned literatures take advantage of the fact that only specific input patterns could cause timing errors, research on imprecise architectures take this one step further by designing simplified circuits for peformance and/or energy efficiency. For example, Lu et~al. proposed a ``shrinking'' datapath that can be utilized to mimic and speculate the original logic functions~\cite{IntelSpeeding}. Kulkarni et.al. described an underdesigned $2\times2$ multiplier unit, of which the worst case was replaced by a normal case based on the straight-forward Karnaugh-Map analysis~\cite{Undersigned2x2multiplier}. In both cases, reduction of area and power consumption can be achieved with the cost of accuracy. In addition, Gupta et~al. developed approximate adders at the transistor level and compared the energy efficiency of their proposed architectures over truncation of input word-length of conventional structures~\cite{Gupta2013TransCADICS}. However, we should note that using imprecise architectures means correct results would never be obtained for certain input patterns. Another limitation is that the link between the probability of output correctness and energy savings or performance improvements are rarely analyzed from current research. In addition, these techniques cannot be directly applied onto existing hardware platforms such as FPGAs.

In this work, we propose an alternative methodology to remove the limitations of those existing approaches. It is argued in this paper that our method can be easily applied onto existing hardware arithmetic operators such as RCAs and CCMs with almost no hardware overhead. We support this hypothesis analytically by presenting theoretical probabilistic error models and experimentally using FPGAs.

\section{Ripple Carry Adder}\label{section_RCA}
\subsection{Adder Structures in FPGAs}
Adders serve as a key building block for arithmetic operations. In general, the ripple carry adder (RCA) is the most straightforward and widely used adder structure. As such, the philosophy of our approach is first exemplified with the analysis of a RCA. We later describe how this methodology can be extended to other arithmetic operators such as CSAs and CCMs, which are commonly used in DSP applicationsn and numerical algorithms, in Section~\ref{Section_CSA} and Section~\ref{section_CCM} respectively.

%and  that are commonly used in DSP applications and numerical algorithms.

Typically the maximum frequency of a RCA is determined by the longest carry propagation delay. Consequently, modern FPGAs offer built-in architectures for very fast ripple carry addition. For instance, the Altera Cyclone series uses fast tables~\cite{AlteraCyclone} while the Xilinx Virtex series employs dedicated multiplexers and encoders for the fast carry logic~\cite{Virtex6}. Fig.~\ref{FPGA adder} illustrates the structure of an $n$-bit RCA, which is composed of $n$ serial-connected full adders (FAs) and utilizes the internal fast carry logic of the Virtex-6 FPGA.

While the fast carry logic reduces the time of each individual carry-propagation delay, the overall delay of carry-propagation will eventually overwhelm the delay of sum generation of each LUT with increasing operand word-lengths. We perform our initial analysis based on the assumption that the carry propagation delay of each FA is a constant value $\mu$, which is a combination of logic delay and routing delay within an FPGA. Hence the critical path delay of the RCA is given by~(\ref{Eq:RCA_delay}), as shown in Fig.~\ref{FPGA adder}.
%
\begin{eqnarray}\label{Eq:RCA_delay}
	\mu_{RCA}=n\cdot\mu
\end{eqnarray}

% $\mu_{RCA}=n\mu$, as shown in~. 
For an $n$-bit RCA, it follows that if the sampling period $T_S$ is greater than $\mu_{RCA}$, correct results will be sampled. If, however, $T_S<\mu_{RCA}$, intermediate results will be sampled, potentially generating errors.

@@

\begin{table*}[htbp]
 %\vspace{-2ex}
 %% increase table row spacing, adjust to taste
   \renewcommand{\arraystretch}{1.05}
   \setlength{\tabcolsep}{3.5pt}
   \tbl{Relative Reduction of MRE ($\%$) in Overclocking Scenario for Various Normalized Frequencies Based on~(\ref{Relative Reduction MRE}) for Two Types of Input Data: the Uniform Data (Uni) and Real Image Data from Lena (Lena).}{
   \centering
   \footnotesize
   %\vspace{-4mm}
   \begin{threeparttable}
   \begin{tabular}{|c|cc|cc|cc|cc|cc||cc|}
     \hline
     \multirow{2}*{\begin{tabular}{c}\textbf{Normalized}\\\textbf{Frequency}\end{tabular}} &
     \multicolumn{2}{c|}{\textbf{FIR}} &
     \multicolumn{2}{c|}{\textbf{Sobel}} &
     \multicolumn{2}{c|}{\textbf{IIR}} &
     \multicolumn{2}{c|}{\textbf{Butterworth}} &
     \multicolumn{2}{c||}{\textbf{DCT4}} &
     \multicolumn{2}{c|}{\textbf{Geo.Mean}}\\
     &  {Uni} & {Lena } &  {Uni} & {Lena } &  {Uni} & {Lena }
     &  {Uni} & {Lena } &  {Uni} & {Lena } &  {Uni} & {Lena }\\
     \hline
     1.04 & 99.85	& 100.00  & 99.51 &	99.74 &	 72.28 &	90.09 &	 79.03 &	 100.00 &	 83.58 &	 98.06 & 86.14 & 97.50\\
     1.08 & 98.93	& 99.97   & 96.26 &	93.75 &	 71.64 &	90.50 &	 78.81 &	 100.00 &	 83.26 &	 98.45 & 85.15 & 96.46\\
     1.12 & 94.27	& 98.82   & 96.25 &	93.62 &	 73.63 &   88.25 &	 81.88 &	 84.87  &	 89.44 &	 99.56 & 86.68 & 92.84\\
     1.16 & 99.66	& 99.91   & 73.73 &	93.62 &	 73.10 &   89.92 &	 79.30 &	 84.23  &	 79.07 &	 99.44 & 80.44 & 93.23\\
     1.20 & 96.03	& 99.90   & 81.55 &	81.52 &	 70.76 &	75.67 &	 64.96 &	 84.50  &	 N/A\tnote{*}	  & N/A\tnote{*} & 77.46 & 84.95 \\
     1.24 & 98.46	& 99.32   & 81.43 &	81.67 &	 70.47 &	76.12 &	 63.66 &	 84.23  &	 N/A\tnote{*}	  & N/A\tnote{*} & 77.44 & 84.92\\
     1.28 & 95.39	& 99.29   & 60.41 &	78.24 &	 N/A\tnote{*}	    &   N/A\tnote{*}     &	54.38 &	 75.15  &	N/A\tnote{*}	  & N/A\tnote{*} & 67.92 & 83.58\\
     1.32 & 95.37	& 98.75   & N/A\tnote{*}	  &	 N/A\tnote{*}	    &	 N/A\tnote{*}	    &	 N/A\tnote{*}   & N/A\tnote{*}   & N/A\tnote{*}    &    N/A\tnote{*}   & N/A\tnote{*} & 95.37	& 98.75 \\
     \hline
   \end{tabular}
   \normalsize
   \begin{tablenotes}
   \footnotesize
    \item[*] Current frequency cannot be achieved in the traditional scenario. These points are excluded from the calculation of geometric means.
   \end{tablenotes}
   \end{threeparttable}}
   \label{Ratio_MRE}
% \vspace{-1ex}
 \end{table*}

\begin{table*}[htbp]
 %\vspace{-2ex}
 %% increase table row spacing, adjust to taste
   \renewcommand{\arraystretch}{1.05}
   \setlength{\tabcolsep}{3.5pt}
   \tbl{Frequency Speedups ($\%$) in Overclocking Scenario Under Various Error Budgets for Two Types of Input Data: Uniform Data (Uni) and Real Image Data from Lena (Lena).}{
%   \vspace{-2ex}
   \label{Exp_Real_Inputs}
   \centering
   \footnotesize
   \begin{tabular}{|c|cc|cc|cc|cc|cc||cc|}
     \hline
     \multirow{2}*{\begin{tabular}{c}\textbf{Error}\\\textbf{Budget($\%$)}\end{tabular}} &
     \multicolumn{2}{c|}{\textbf{FIR}} &
     \multicolumn{2}{c|}{\textbf{Sobel}} &
     \multicolumn{2}{c|}{\textbf{IIR}} &
     \multicolumn{2}{c|}{\textbf{Butterworth}} &
     \multicolumn{2}{c||}{\textbf{DCT4}} &
     \multicolumn{2}{c|}{\textbf{Geo.Mean}} \\
     %\multirow{2}*{\begin{tabular}{c}\textbf{Geo.}\\\textbf{Mean}\end{tabular}}\\
     & {Uni} & {Lena } &  {Uni} & {Lena } &  {Uni} & {Lena }
     & {Uni} & {Lena } &  {Uni} & {Lena } &  {Uni} & {Lena }\\
     \hline
     0.05 & 4.76 &	21.43&	 6.82&	6.82&	0.95&	 1.26&	12.40&	 24.03&	 0.72&	 0.96  &3.07 & 5.32\\
     0.5  & 19.05&	21.43&	 13.64&	6.82&	0.63&	 10.06&	24.03&	 24.03&	0.48&	 12.44   &4.52 &13.45\\
     1    & 19.05&	28.57&	 13.64&	18.18&	 10.06&	16.35&	 24.03&	 24.03&	 0.48&	 12.44    &7.86 &19.10\\
     5    & 19.15&	25.53&	 18.18&	18.18&	 0.54&	0.82&	 0.63&	 0.94&	 7.66&	 12.44    &3.91    &5.36\\
     10   & 19.15&	25.53&	 10.64&	10.64&	 0.54&	1.09&	 6.92&	 13.21&	 4.91&	 4.911  &5.19  &7.19\\
     20   & 19.15&	25.53&	 5.77&	15.39&	 8.70&	8.70&	 3.26&	 3.26&	 6.70&	 10.88    &7.32    &10.39\\
     50   & 15.69&	15.69&	 10.53&	19.30&	 42.50&	50.00&	 46.74&	 54.89&	 15.06&	 19.25    &21.8 &27.59\\
     \hline
   \end{tabular}}
   \normalsize
   \label{Max_Frequency}
%   \vspace{-3ex}
 \end{table*}
